Ideal model selection is impossible

### 2014-11-24 Facticity/Facticity/paper.tex
reshuffle

@@

\begin{definition}[Enumerations]
  Let $\phi_1,\phi_2,\ldots$ be a fixed acceptable enumeration of all partial
  recursive functions, e.g. the enumeration described in \cite{TODO}.  
\end{definition}

\begin{definition}[Representations]
  A pair $(m,y)$ is a \emph{representation} for $x$ if $m$ is a model
  and $m(y)=x$. Given a model class $\M$, define the set of all
  representations for $x$:
  \[
  R^\M(x):=\{(m,y):m\in\M, m(y)=x\}.
  \]
\end{definition}

@@

\begin{definition}[Candidate set]
  The \emph{candidate models} for $x$ from model class $\M$ are
  \[
  C^\M_c(x):=\{m\in\M:L^m(x)\le L^\M(x)+c\}.
  \]
\end{definition}

\begin{definition}[Sophistication]
  The model complexity, sophistication, or facticity, of a string $x$
  given a model class $\M$ and a constant $c$ is defined by
  \[
  S^\M_c(x):=\min\{L(m):m\in C^\M_c(x)\}.
  \]
\end{definition}

\paragraph{Kolmogorov's Structure Function}

@@

\paragraph{Koppel Sophistication}

\begin{itemize}
\item Model class is $\T$
\item Description method for models is required to be prefix-free, but
  can otherwise be chosen arbitrarily.
\item Koppel uses monotone complexity and discusses infinite strings,
  TODO figure out if and how this affects our results
\end{itemize}

\paragraph{Vit\'anyi's Sophistication / Meaningful Information}

@@

\paragraph{Antunes's Coarse Sophistication}

\begin{itemize}
\item Model class is $\T$
\item Description method for models is unclear, but appears to be
  taken from Koppel, i.e. arbitrary but self-delimiting.
\item Candidate set is defined differently:
  $C^\T(x)=\{m^*\}$ where $m^*=\argmin_m L(m)+L^m(x)-C(x)$. Thus, there is no slack
  variable and $C^\T$ contains only a single model.
\end{itemize}

\paragraph{Adriaans' Facticity}

\begin{itemize}
\item Model class is $\C$
\item Description method for models is $L(m)=|\bar\imath|$ where
  $\phi_i=m$. The index function is assumed to be \emph{faithful},
  i.e. there is a constant $c$ such that $|i|-C(i)\le c|$ for all $i$.
\item Candidate set is $C^\C_0$.
\item Facticity is actually defined as index length without prefix
  encoding:  $S^\C_0(x):=\min\{|i|:\phi_i\in C^\C_0(x)\}$. But this
  difference is not important for the discussion.
\end{itemize}

@@

\begin{itemize}
\item Paul argues that the randomness deficiency of a model is a measure of the amount of structure left in the noise part. I feel that it is also a measure of the amount of counterevidence against a model in the statistical sense, as follows: the randomness deficiency of $x$ given model $M$ is $\delta(x|M)=L_M(x)-K(x|M)$. Suppose that model $M$ is true. We can then use the no hypercompression inequality to rewrite:
%
\[
P_{x\sim M}(\delta(x|M)\ge c) = P_{x\sim M}(L_M(x) - K(x|M) \ge c)\le 2^{-c}
\]
Thus, if $M$ is true the probability of a large randomness deficiency is exponentially small! This amount of evidence is robust in the sense that changing the UTM around will change the amount of evidence by at most a constant. If the structure function is going to give us a \emph{cloud} of candidate models instead of just a single one, I think this is a really nice interpretation of the meaning of that cloud: we simply measure the available evidence against all individual models.

@@

@@

\begin{lemma}[Framework equivalence]
Let $\phi_1,\phi_2,\ldots$ be a compact effective enumeration. There is a constant $c$ such that for any index $i$ there is another index $j$ with $\phi_j(x)=\phi_i(x)$ for all $x$, and
\[|j|\le K(i)+2\log(K(i))+c.\]
\end{lemma}
\begin{proof}
First, let $v$ be the index of the partial recursive function given by $\phi_v(\bar k\bar p x)=\phi_{\phi_k(p)}(x)$. Let $\bar k p$ be a shortest program for $i$, that is $|\bar k p| = K(i)$ and $\phi_k(p)=i$. Then $\phi_i(x)=\phi_v(\bar k \bar p x)$. We now use compactness to find an index $j$ with $\phi_v(\bar k \bar p x)=\phi_j(x)$ and length
\[|j|\le |v\bar k\bar p|\le|v|+|\bar k p|+2\log|\bar k p|=K(i)+|v|+2\log K(i),\]
as required.
\end{proof}

We need a definition of $K(f)$. We adopt the definition given in \cite{TODO}:

\begin{definition}
  The Kolmogorov complexity of a partial recursive function is defined as
  \[
  K(f) = \min\{K(i):\forall_x \phi_i(x)=f(x).\}
  \]
\end{definition}

It is not immediately clear that this is a reasonable definition of the shortest effective description of a function, but fortunately this is indeed the case:

\begin{lemma}[TODO: $i$ moet prefix encoded zijn, anders klopt t niet. Ook beetje moeilijk te begrijpen]
Let $\psi_1,\psi_2,\ldots$ be any acceptable numbering. There is a constant $c$ such that for any $i$ with $\psi_i(x)=f(x)$ for all $x$, the description length satisfies $K^\phi(f)\le |i|+c$.
\end{lemma}
\begin{proof}
The numbering is acceptable, so there is a total computable function $\phi_r$ that maps indices in the $\psi$-enumeration to indices in the $\phi$-enumeration. Let $j=\phi_r(i)$. Then $\bar r i$ is a program for $j$, so $K(j)\le |\bar r i| = |i|+c$. Now,
\[
K(f)=\min\{K(k):\forall_x\phi_k(x)=f(x)\}\le K(j)\le|i|+c
\]
as required.
\end{proof}

\begin{lemma}[Invariance]
Let $\phi_1, \phi_2, \ldots$ and $\psi_1, \psi_2,\ldots$ be two acceptable numberings. There exists a constant $c$ such that $\left| K^\phi(f) - K^\psi(f)\right | \leq c$ for all $f$. \label{lemma:invariance}
\end{lemma}
\begin{proof}
Let $g(i)$ be the function which which translates from $\psi$ to $\phi$.
\begin{align*}
K^\phi(f) &= \min\left\{ K^\phi(i) : \phi_i= f\right\} \\
&\geq \min\left\{ K^\psi(i) : \phi_i= f\right\} - c\\
&= \min\left\{ K^\psi(i) : \psi_{g(i)}= f\right\} - c\\
&\geq \min\left\{ K^\psi(i) : \psi_i= f\right\} - c\\
= K^\psi(f)
\end{align*}
We can reverse $\phi$ and $\psi$ without loss of generality to achieve the same inequality the other way around, completing the proof.
\end{proof}

\begin{lemma}[TODO: is het niet leuker om te zeggen dat het altijd bounded is? Krijgen we dat kado van een later lemma?]
There is an acceptable numbering $\psi_1,\psi_2,\ldots$ such that $F^\psi(x)=|\bar 1|$ for all $x$.
\end{lemma}
\begin{proof}
Let $\phi_1,\phi_2,\ldots$ be another acceptable numbering given by
\[
\phi_i(x)=\begin{cases}
u&\tn{if $i=1$}\\
\infty&\tn{if $1<i<r$}\\
\psi_{i-r+1}(x)&\tn{if $i\ge r$}.
\end{cases}\]
Here $u\ge r$ is the index of the universal element satisfying $\phi_u(\bar\jmath p)=\psi_j(p)$.
Now the two-part codelength of $x$ using representation $\phi_u$ is
\[\begin{split}
K(\phi_u)+\min\{|y|:\phi_u(y)=x\}&=\min\{K(i):\forall_x\phi_i(x)=\phi_u(x)\}+\min\{|y|:\phi_u(y)=x\}\\
&=|\bar 1|+\min\{|\bar\jmath p|:\phi_u(\bar\jmath p)=x\}\\
&=|\bar 1|+\min\{|\bar\jmath p|:\psi_j(p)=x\}\\
&=|\bar 1|+K^\psi(x).
\end{split}\]
On the other hand, the two-part codelength of $x$ using a representation $s\ne u$ is
\[\begin{split}
K(\phi_s)+\min\{|y|:\phi_s(y)=x\}&=\min\{K(i):\forall_x\phi_i(x)=\phi_s(x)\}+\min\{|y|:\phi_s(y)=x\}\\
&=c+\min\{|y|:\psi_{s-r+1}(y)=x\}\\
\end{split}\]

\end{proof}

\begin{definition}[Facticity]
  Let $\phi_1,\phi_2,\ldots$ be any effective enumeration of the partial recursive functions. Then
  \[F^\phi(x)=|\bar\imath|\qquad\tn{where}\qquad \bar\imath p=\min\{\bar\jmath q:\phi_j(q)=x\}.
  \]
  Here the minimum is taken with respect to the standard ordering of binary sequences.
\end{definition}

\begin{definition}[Model complexity]
Given a model class $\M$, the model complexity of a data sequence $x$ is given by
\[G(x)=K(f^*)\qquad\tn{where}\qquad (f^*,p^*)=\argmin_{f,p:f\in\M, f(p)=x}(K(f)+|p|),
\]
where ties are broken by picking a model $f^*$ of minimal complexity.
\end{definition}

\begin{theorem}
There are enumerations $\phi_1,\phi_2,\ldots$ and $\psi_1,\psi_2,\ldots$ of the partial recursive (prefix) functions such that for every constant $c$ there is a binary sequence $x$ with
\[
|G^\phi(x)-G^\psi(x)|>c
\]
\end{theorem}

\begin{lemma}
There is an acceptable numbering $\phi_1,\phi_2,\ldots$ of all partial recursive prefix functions, and a constant $c$ such that for all $n$ we have $G^\phi(x)<c$ for more than half of all strings $x$ of length $n$ with respect to the model class of all partial recursive prefix functions.
\end{lemma}

\begin{lemma}
There is an enumeration $\phi_1,\phi_2,\ldots$ of the partial recursive prefix functions such that for every constant $c$ there is an $n_0$ such that for all $n>n_0$ we have $G^\phi(x)>c$ for more than half of all strings $x$ of length $n$ provided the model class contains programs for every singleton $x$.
\end{lemma}

\begin{theorem}
There are enumerations $\phi_1,\phi_2,\ldots$ and $\psi_1,\psi_2,\ldots$ of the partial recursive prefix functions such that for every constant $c$ there is a binary sequence $x$ with
\[
|G^\phi(x)-G^\psi(x)|>c
\]
with respect to the model class of all partial recursive prefix functions.
\end{theorem}

\begin{definition}
Let $\psi_1,\psi_2,\ldots$ be an arbitrary acceptable numbering of the partial recursive functions. Define
\[
B^n = \{ x : |x|=n~\tn{and}~n-c \le K^\psi(x) \le n+c \}
\]
\end{definition}

\begin{lemma}
  For sufficiently large $c$ the sets $B^n$ are nonempty for all $n$.
\end{lemma}

\begin{lemma}[Print function]
  Let $\M$ be any model class that contains the identity function. There is an acceptable numbering $\phi_1,\phi_2,\ldots$ of the partial recursive functions such that for every $n$, all $x\in B^n$, 
we have $G^\phi_M(x)=1$.
\end{lemma}

\section{Unbounded and top-heavy sophistications}

\begin{definition}
We call a numbering $\eta_1,\eta_2,\ldots$ \emph{optimal} if there is a constant $c$ such that for all $i$, we have $\min\{|j|:\eta_i=\eta_j\} \leq K(\eta_i) + c$.
\end{definition}

\begin{lemma}
There exists an acceptable, optimal numbering of the partial computable functions.
\label{lemma:optimal-numbering}
\end{lemma}
\begin{proof}
Let $\psi_1, \psi_2, \ldots$ be some acceptable numbering, with some $\psi_{\text{div}}$ with is undefined for all its arguments. Let 
\[\begin{cases}
	\phi_{j} = \psi_{\text{div}} &\text{if $j$ can not be read as $\bar\imath p$}\\
	\phi_{\bar\imath p} = \psi_{\psi_i(p)} &\text{if $\psi_i(p)$ is defined}\\
	\phi_{\bar\imath p} = \psi_{\text{div}} & \text{otherwise}
\end{cases}\]

\noindent To show that this numbering is acceptable, we must provide two computable functions to translate to and from a given acceptable numbering. We use $\psi$ for this purpose. To find a $\phi_j$ which corresponds to a given $\psi_i$, we can simply use the identity function $\psi_d(x) = x$. For a given $\psi_i$, the function $\phi_{\overline{d}i}$ computes the same function.

To translate from $\phi_j$ to $\phi_i$, we are faced with a problem. If we are given $j = \bar\imath p$, we cannot decide algorithmically whether $\psi_i(p)$ is defined. However, we can construct the function $\psi_r(\bar\imath p, q)$ such that it computes $\psi_r{\psi_i(p)}(q)$ if $\psi_i(p)$ is defined and is undefined otherwise, or if the first argument cannot be read as $\bar\imath p$. For a given $j$ we can also find an $r'$ such that $\psi_{r'}(q) = \psi_(j, q)$. Thus, given any $\phi_i$ we can construct and return this function.

Finally, we'll show that the minimum length of an index for function $f$ in $\{\phi_i\}$ is equal to $K(f)$ up to a constant:
\begin{align*}
K^{\psi}(f) &= \min\left\{K^{\psi}(j)\,:\,\psi_j = f \right\} \\
&= \min_j\left\{\min_{i, p} \{|\bar\imath p| \,:\,\psi_i(p) = j\} \,:\,\psi_j = f\right\} \\
&= \min_{i, p} \left\{|\bar\imath p| \,:\,\psi_i(p) = j, \psi_j = f\right\} \\
&= \min_j \left\{|j|\,:\, \phi_j = f \right\}
\end{align*}
\end{proof}

\begin{lemma}
Let $M$ be the class of \emph{singleton functions}, that is, the functions which produce a single output for all arguments, including the function which is undefined for all its arguments. There exists an optimal partial computable numbering of $M$.
\end{lemma}
\begin{proof}
Let $\psi_1, \psi_2, \ldots$ be some acceptable numbering. Let $\phi_{\bar\imath p}(\cdot) = \psi_i(p)$ \ldots. 
\end{proof}

\begin{definition}
  Given a model $f$ and model class $\M$, define model-based complexities by
  \begin{align*}
  C^f(x) &= \min\{|y|:f(y)=x\} \text{,}\\
  L^f(x) &= K(f)+C^f(x) \text{,}\\
  L^{\M}(x) &= \min\{L^f(x)\mid f\in\M\}\text{.}
  \end{align*}
\end{definition}

Plan: een lemma voor total/niet prefix dat laat zien mbv pigeonhole dat facticity juist bounded of juist unbounded kan zijn. De ene kant op gaat dat met de print functie (klein model, grote data). De andere kant op is moeilijk

>>>>

All numberings should be acceptable!

@@

\begin{definition}[Complexity]
  Let $\phi_1,\phi_2,\ldots$ be a fixed acceptable enumeration of all partial
  recursive functions, e.g. the enumeration described in \cite{TODO}.
  Let $\Pi$ be a computable subset of the natural numbers such that
  $\{\phi_i:i\in\Pi\}$ contains all computable prefix functions. Now
  define Kolmogorov complexity by
\begin{align*}
C(x)&=\min\{|\bar\imath| y:\phi_i(y)=x\},\\
K(x)&=\min\{|\bar\imath| y:\phi_i(y)=x, i\in\Pi\}.\\
\end{align*}
\end{definition}
This definition of $K$ is equivalent to the definition in the
textbook~\cite{TODO}, but it is more convenient for defining the
$K$-complexity of non-prefix functions:

\begin{definition}[Representations]
  A pair $(m,y)$ is a \emph{representation} for $x$ if $m$ is a model
  and $m(y)=x$.
\end{definition}

@@

\begin{definition}[Candidate set]
  The \emph{candidate models} for $x$ from model class $\M$ and
  threshold function $\tau$ are
  \[
  C^\M_\tau(x):=\{m\in\M:L^m(x)\le \tau(x)\}.
  \]
\end{definition}

\begin{definition}[Sophistication]
  The model complexity, sophistication, or facticity, of a string $x$
  given a model class $\M$ and threshold function $\tau$ is defined by
  \[
  S^\M_\tau(x):=\min\{L(m):m\in C^\M_\tau(x)\}.
  \]
\end{definition}

\section{Measures of meaningful information}

\subsection{Koppel Sophistication}

\begin{itemize}
\item Model class is $\T$
\item Threshold function $\tau(x)=L^\M(x)+c$
\item Description method for models is required to be prefix-free, but
  can otherwise be chosen arbitrarily.
\item Koppel uses monotone complexity and discusses infinite strings,
  TODO figure out if and how this affects our results
\end{itemize}

This does not work because of the nickname problem! To fix this:

\begin{definition}
  The complexity of a partial recursive function is defined by
  \begin{align*}
    C(f) &:= \min\{C(i):\phi_i=f\}\\
    K(f) &:= \min\{K(i):\phi_i=f\}.
  \end{align*}
\end{definition}

These definitions make sense because:

\begin{lemma}[Invariance]
Let $\phi_1, \phi_2, \ldots$ and $\psi_1, \psi_2,\ldots$ be two acceptable numberings. There exists a constant $c$ such that $\left| K^\phi(f) - K^\psi(f)\right | \leq c$ for all $f$. \label{lemma:invariance}
\end{lemma}
\begin{proof}
Let $g(i)$ be the function which which translates from $\psi$ to $\phi$.
\begin{align*}
K^\phi(f) &= \min\left\{ K^\phi(i) : \phi_i= f\right\} \\
&\geq \min\left\{ K^\psi(i) : \phi_i= f\right\} - c\\
&= \min\left\{ K^\psi(i) : \psi_{g(i)}= f\right\} - c\\
&\geq \min\left\{ K^\psi(i) : \psi_i= f\right\} - c\\
&= K^\psi(f).
\end{align*}
We can reverse $\phi$ and $\psi$ without loss of generality to achieve the same inequality the other way around, completing the proof.
\end{proof}

Thus, the complexity of a function depends on the choice of
enumeration by no more than a constant term. Note that this implies
that for any enumeration $\psi_1,\psi_2,\ldots$, the complexity $K(\psi_i)$
is less than the literal description length of the index
$|\bar\imath|$ up to a constant. For some numberings, there is a gap
between these two description methods; in those cases, the index is an
inefficient representation of the corresponding function. However it
is possible to construct numberings that are \emph{faithful} in the
sense that no such gap exists:

\begin{definition}[Faithful Numbering]\label{def:faithful}
  A numbering $\psi_1,\psi_2,\ldots$ of the partial recursive
  functions is \emph{faithful} if there is a constant $c$ such that
  for all indices $i$ there is a $j$ such that $\psi_i=\psi_j$ and
  $|j|\le C(\psi_j)+c$.
\end{definition}

\begin{lemma}
  There is a faithful acceptable numbering.
\end{lemma}
\begin{proof}
Let $i_\tn{div}(y)$ be an index such that $\phi_{i_\tn{div}}(y)=\infty$ for all $y$. Define
  \[\psi_q=\begin{cases}
    \phi_{\phi_i(p)}&\tn{if $q$ can be written as $\bar\imath p$ and $\phi_i(p)<\infty$,}\\
    \phi_{i_\tn{div}}&\tn{otherwise.}\end{cases}
  \]
  To show that $\psi$ is faithful, pick any function $f$. Then
\[\begin{split}
C(f)&=\min\{C(i):\phi_i=f\}\\
&=\min\{\min\{|\bar a b|:\phi_a(b)=i\}:\phi_i=f\}\\
&=\min\{|\bar a b|:\phi_{\phi_a(b)}=f\}\\
&=\min\{|\bar a b|:\psi_{\bar a b}=f\}.
\end{split}\]
This shows there is a sufficiently small $\psi$ index.

To show that $\psi$ is acceptable, let $\phi_j$ denote the identity
function. Then a $\phi$-index $i$ can be mapped to a $\psi$-index
using the computable function $r(i)=\bar\jmath i$, so that
$\psi_{r(i)}(y)=\psi_{\bar\jmath i}(y)=\phi_i(y)$. For the reverse,
define $\phi_v(\bar\imath p, y)=\phi_{\phi_i(p)}(y)$. For fixed
$\bar\imath p$, the 
$s^n_m$-theorem (see~\cite{TODO}) states that we can compute the $h$
such that $\phi_h(y)=\phi_v(\bar\imath p,y)$. Let $h(\bar\imath p)$
denote this index as a function of the program; further define
$h(q)=i_\tn{div}$ if $q$ cannot be expressed as $\bar\imath p$. By
construction $h$ is total and computable. To check that the mapping
returns the correct function, rewrite $\phi_{h(\bar\imath
  p)}(y)=\phi_v(\bar\imath p,y)=\phi_{\phi_i(p)}(y)=\psi_{\bar\imath p}(y)$.
\end{proof}

\subsection{Antunes's Coarse Sophistication}

\begin{itemize}
\item Model class is $\T$
\item Description method $L(m)$ for models is unclear, but appears to be
  taken from Koppel, i.e. arbitrary but self-delimiting.
\item Sophistication is defined slightly differently: $S^\T(x)=L(m^*)$ where $m^*=\argmin_m L(m)+L^m(x)-C(x)$. Thus, there is no slack.
\end{itemize}

\subsection{Adriaans' Facticity}

\begin{itemize}
\item Model class is $\C$
\item Description method for models is $L(m)=|\bar\imath|$ where
  $\phi_i=m$. The index function is assumed to be \emph{faithful}, see
  Definition~\ref{def:faithful}
\item Threshold function is $\tau(x)=L^m(x)$.
\item Facticity is actually defined as index length without prefix
  encoding:  $S^\C_0(x):=\min\{|i|:\phi_i\in C^\C_0(x)\}$. But this
  difference is not important for the discussion.
\end{itemize}

\subsection{Roll your own}

\begin{itemize}
\item Model class is either $\C$ or $\K$
\item Description method for models is $L(m)=K(m)$.
\item Threshold function is $\tau(x)=L^m(x)+c$, which matches $C(x)$
  resp. $K(x)$ up to a constant.
\end{itemize}

\paragraph{TODO: instantiate for UTM}

\begin{itemize}
\item Model class is $\K$
\item Description method for models is $L(m)=K(m)$.
\item Threshold function is $\tau(x)=L^m(x)+c$, which matches $K(x)$
  up to a constant.
\end{itemize}

\subsection{Kolmogorov's Structure Function}

@@

\subsection{Vit\'anyi's Sophistication / Meaningful Information}

@@

% stru fu: model class is all finite sets
% code length L(m)+L_m(x)
% acceptability criterion acc(m) = 2L(m)+L_m(x)-C(x)
%

@@

\begin{itemize}
\item Paul argues bleurg that the randomness deficiency of a model is a measure of the amount of structure left in the noise part. I feel that it is also a measure of the amount of counterevidence against a model in the statistical sense, as follows: the randomness deficiency of $x$ given model $M$ is $\delta(x|M)=L_M(x)-K(x|M)$. Suppose that model $M$ is true. We can then use the no hypercompression inequality to rewrite:
%
\[
P_{x\sim M}(\delta(x|M)\ge c) = P_{x\sim M}(L_M(x) - K(x|M) \ge c)\le 2^{-c}
\]
Thus, if $M$ is true the probability of a large randomness deficiency is exponentially small! This amount of evidence is robust in the sense that changing the UTM around will change the amount of evidence by at most a constant. If the structure function is going to give us a \emph{cloud} of candidate models instead of just a single one, I think this is a really nice interpretation of the meaning of that cloud: we simply measure the available evidence against all individual models.

@@

\appendix
\section{Dead darlings}

@@

\begin{lemma}
Compact numberings are sort of faithful.
\end{lemma}
\begin{proof}
First, let $v$ be the index of the partial recursive function given by $\phi_v(\bar k p x)=\phi_{\phi_k(p)}(x)$. Let $\bar k p$ be a shortest program for $i$, that is $|\bar k p| = K(i)$ and $\phi_k(p)=i$. Then $\phi_i(x)=\phi_v(\bar k \bar p x)$. We now use compactness to find an index $j$ with $\phi_v(\bar k \bar p x)=\phi_j(x)$ and length
\[|j|\le |v\bar k\bar p|\le|v|+|\bar k p|+2\log|\bar k p|=K(i)+|v|+2\log K(i),\]
as required.
\end{proof}
